Clustering exploration for `r params$sample_id`

### 2024-09-12 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
add to PR #750

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE
                      )
```

@@

The aim is to explore the clustering and label transfers for the sample `r params$sample_id`.

@@

```{r packages, message=FALSE, warning=FALSE}
library("Seurat")
library(sctransform)

library(SCpubr)             # for plotting 
library(tidyverse)
library(patchwork)
library(msigdbr)
library(enrichplot)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggalluvial)
library(ggsankey)
set.seed(params$seed)
```

@@

In this notebook, we are working with the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.
We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.

@@

#### do_SankeyPlot

`do_SankeyPlot` is largely inspired by the source code from `SCpubr` as the function has been removed from the package.
https://github.com/enblacar/SCpubr/blob/v1.1.2-dev-stable/R/do_SankeyPlot.R

- `first_group` is a metadata variable. First group of nodes of the sankey plot.
- `last_group` is a metadata variable. Last group of nodes of the sankey plot.
- `middle_groups` is a metadata variable. Vector of groups of nodes of the sankey plot.
- `type` is a type of plot to make. One of:

  - `sankey`: Generates a sankey plot.
  - `alluvial`: Generated an Alluvial plot, a kind of sankey plot where all groups have the same height.

- `width` Width of the nodes.
- `space` Vertical space between the nodes. It appears to be equal to a single cell. Use big numbers to see a difference (like, 1000 or 10000).
- `position` GGplot2 position.
- `node.fill` Color to fill the nodes.
- `node.color` Color for the contour of the nodes.
- `flow.alpha` Alpha of the connections.
- `flow.color` Color for the contour of the connections.
- `text_size` Size of the labels.
- `text_color` Color of the labels.
- `smooth` How smooth the connections are.
- `use_labels` Whether to use labels or text for the node names.
- `hjust` General hjust for the labels.

```{r }

do_SankeyPlot <- function(sample,
                          first_group,
                          last_group,
                          type = "sankey",
                          middle_groups = NULL,
                          width = 0.1,
                          space = ifelse(type == "sankey", 0.05 * ncol(sample), 0),
                          position = "identity",
                          node.fill = "white",
                          node.color = "white",
                          flow.alpha = 0.75,
                          flow.color = "black",
                          text_size = 3,
                          text_color = "black",
                          font.size = 18,
                          font.type = "sans",
                          smooth = 8,
                          use_labels = FALSE,
                          hjust = NULL,
                          colors.first = NULL,
                          colors.middle = NULL,
                          colors.last = NULL,
                          plot.title = NULL,
                          plot.subtitle = NULL,
                          plot.caption = NULL){

  # Check logical parameters.
  logical_list <- list("use_labels" = use_labels)
  # Check numeric parameters.
  numeric_list <- list("width" = width,
                       "space" = space,
                       "flow.alpha" = flow.alpha,
                       "text_size" = text_size,
                       "font.size" = font.size,
                       "smooth" = smooth,
                       "hjust" = hjust)
  # Check character parameters.
  character_list <- list("first_group" = first_group,
                         "last_group" = last_group,
                         "middle_groups" = middle_groups,
                         "type" = type,
                         "position" = position,
                         "node.color" = node.color,
                         "flow.color" = flow.color,
                         "text_color" = text_color,
                         "font.type" = font.type,
                         "colors.first" = colors.first,
                         "colors.middle" = colors.middle,
                         "colors.last" = colors.last,
                         "node.fill" = node.fill,
                         "plot.title" = plot.title,
                         "plot.subtitle" = plot.subtitle,
                         "plot.caption" = plot.caption)

  `%>%` <- magrittr::`%>%`

  data <- suppressWarnings({sample@meta.data %>%
                            dplyr::select(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
                            tibble::rownames_to_column(var = "cell") %>%
                            dplyr::select(-.data$cell) %>%
                            ggsankey::make_long(dplyr::all_of(c(first_group, middle_groups, last_group))) %>%
                            dplyr::rowwise() %>%
                            dplyr::mutate(hjust = if(.data$x %in% middle_groups){0.5}
                                          else if (.data$x == last_group){0}
                                          else if (.data$x == first_group){1})})
  if (!is.null(hjust)){data$hjust <- hjust}

  colors.first <- SCpubr::do_ColorPalette(colors.use = "steelblue",
                        n = length(unique(sample@meta.data[, first_group])))
    if (is.factor(sample@meta.data[, first_group])){
      names(colors.first) <- levels(sample@meta.data[, first_group])
    } else {
      names(colors.first) <- sort(unique(sample@meta.data[, first_group]))
    }

  colors.last <- SCpubr::do_ColorPalette(colors.use = "steelblue",
                        n = length(unique(sample@meta.data[, last_group])))
    if (is.factor(sample@meta.data[, last_group])){
      names(colors.last) <- levels(sample@meta.data[, last_group])
    } else {
      names(colors.last) <- sort(unique(sample@meta.data[, last_group]))
    }

  colors.use <- c(colors.first, colors.last)
  func_use <- ifelse(isTRUE(use_labels), ggsankey::geom_sankey_label, ggsankey::geom_sankey_text)

  p <- data %>%

       ggplot2::ggplot(mapping = ggplot2::aes(x = .data$x,
                                              next_x = .data$next_x,
                                              node = .data$node,
                                              next_node = .data$next_node,
                                              fill = factor(.data$node),
                                              label = .data$node,
                                              hjust = .data$hjust)) +
       ggsankey::geom_sankey(flow.alpha = flow.alpha,
                             node.color = node.color,
                             node.fill = node.fill,
                             color = flow.color,
                             width = width,
                             position = position,
                             type = type,
                             space = space) +
       func_use(size = text_size,
                color = text_color,
                fontface = "bold",
                position = position,
                type = type,
                space = space) +
       ggplot2::scale_fill_manual(values = colors.use) +
       ggplot2::xlab("") +
       ggplot2::ylab("") +
       ggplot2::labs(title = plot.title,
                     subtitle = plot.subtitle,
                     caption = plot.caption) +
       ggplot2::theme_minimal(base_size = font.size) +
       ggplot2::theme(axis.title = ggplot2::element_text(color = "black",
                                                         face = "bold"),
                      axis.line.x = ggplot2::element_blank(),
                      axis.text.x = ggplot2::element_text(color = "black",
                                                          face = "bold",
                                                          angle = 0,
                                                          hjust = 0.5,
                                                          vjust = 1),
                      axis.text.x.top = ggplot2::element_text(color = "black",
                                                              face = "bold",
                                                              angle = 0,
                                                              hjust = 0.5,
                                                              vjust = 1),
                      axis.text.y = ggplot2::element_blank(),
                      axis.ticks = ggplot2::element_blank(),
                      panel.grid.major = ggplot2::element_blank(),
                      plot.title.position = "plot",
                      plot.title = ggplot2::element_text(face = "bold", hjust = 0),
                      plot.subtitle = ggplot2::element_text(hjust = 0),
                      plot.caption = ggplot2::element_text(hjust = 1),
                      panel.grid = ggplot2::element_blank(),
                      text = ggplot2::element_text(family = font.type),
                      plot.caption.position = "plot",
                      legend.text = ggplot2::element_text(face = "bold"),
                      legend.position = "none",
                      legend.title = ggplot2::element_text(face = "bold"),
                      legend.justification = "center",
                      plot.margin = ggplot2::margin(t = 10, r = 10, b = 10, l = 10),
                      plot.background = ggplot2::element_rect(fill = "white", color = "white"),
                      panel.background = ggplot2::element_rect(fill = "white", color = "white"),
                      legend.background = ggplot2::element_rect(fill = "white", color = "white"),
                      strip.text =ggplot2::element_text(color = "black", face = "bold"))

@@

`do_Table_Heatmap` shows heatmap of counts of cells in groups of metadata

@@

  df <- data@meta.data %>%
  mutate_if(sapply(data@meta.data, is.character), as.factor)  %>% 
  group_by( !!sym(first_group), !!sym(last_group))%>% 
  summarise(Nb = n())

p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20))

@@

```{r fig.width=20, fig.height=15, out.width='100%'}

@@

p2 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =  "fetal_full_predicted.annotation.l1",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "fetal_full_predicted.annotation.l1"
                          )

p3 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =   "seurat_clusters",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title =  "seurat_clusters"
                          )

p4 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =   "cellassign_celltype_annotation",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "cellassign_celltype_annotation"
                          )
(p1|p2)/(p3|p4)

@@

#### Sankey plots

```{r fig.width=20, fig.height=20, out.width='100%'}

p1 <- do_SankeyPlot(sample = srat,
                            first_group = "fetal_kidney_predicted.compartment",
                            last_group = "seurat_clusters",
                             type = "sankey",text_size = 5, use_labels = TRUE
                            )

p2 <- do_SankeyPlot(sample = srat,
                            first_group = "fetal_full_predicted.annotation.l1",
                            last_group = "seurat_clusters",
                             type = "sankey",text_size = 5, use_labels = TRUE
                            )

p3 <- do_SankeyPlot(sample = srat,
                            first_group = "singler_celltype_annotation",
                            last_group = "seurat_clusters",
                             type = "sankey",text_size = 4, use_labels = TRUE
                            )

p4 <- do_SankeyPlot(sample = srat,
                            first_group = "cellassign_celltype_annotation",
                            last_group = "seurat_clusters",
                             type = "sankey",text_size = 5, use_labels = TRUE
                            )

```{r fig.width=30, fig.height=10, out.width='100%'}
p1 <- do_SankeyPlot(sample = srat,
                            first_group = "fetal_kidney_predicted.compartment",
                            last_group = "fetal_full_predicted.annotation.l1",
                            type = "sankey",text_size = 6, use_labels = TRUE
 )

p2 <- do_SankeyPlot(sample = srat,
                            first_group = "fetal_kidney_predicted.compartment",
                            last_group = "fetal_kidney_predicted.cell_type",
                            type = "sankey",text_size = 6, use_labels = TRUE                                                        )

p1 | p2
```

#### Heatmaps

@@

>>>>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE
                      )

@@

```{r}

subdiagnosis <- readr::read_tsv(
  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
  show_col_types = FALSE
  ) |>
  dplyr::filter(scpca_sample_id == params$sample_id) |>
  dplyr::pull(subdiagnosis)

The aim is to explore the clustering and label transfers for the sample `r params$sample_id`.
This sample has a(n) `r subdiagnosis` subdiagnosis.

@@

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(sctransform)
library(SCpubr)             # for plotting 
library(tidyverse)
library(patchwork)
library(msigdbr)
library(enrichplot)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggalluvial)

set.seed(params$seed)
```

@@

In this notebook, we are working with the Wilms tumor sample defined in `r params$sample_id` from the Wilms tumor dataset `r params$scpca_project_id`.
We work with the pre-processed and labeled `Seurat` object that is the output of `02b_label-transfer_fetal_kidney_reference_Stewart.Rmd` saved in the `results` directory.

@@

@@

`do_Table_Heatmap` shows heatmap of counts of cells for combinations of two metadata variables

@@

  df <- data@meta.data %>%
  mutate_if(sapply(data@meta.data, is.character), as.factor)  %>% 
  group_by( !!sym(first_group), !!sym(last_group))%>% 
  summarise(Nb = log(n()))

p <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Number of \n cells (log)"))

@@

Here we evaluate with marker genes the identification of endothelial, immune cells and other cancer and non-cancerous sub-populations.

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}

@@

p2 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =  "fetal_kidney_predicted.cell_type",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "fetal_kidney_predicted.cell_type"
                          )

p3 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =   "fetal_full_predicted.annotation.l1",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title =  "fetal_full_predicted.annotation.l1"
                          )

p4 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =  "fetal_full_predicted.annotation.l2",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "fetal_full_predicted.annotation.l2"
                          )

p5 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =   "seurat_clusters",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title =  "seurat_clusters"
                          )

p6 <- SCpubr::do_DotPlot(sample = srat,
                         group.by =  "cellassign_celltype_annotation",
                         features = markers, 
                         axis.text.x.angle = 90, 
                         flip = TRUE,
                         plot.title = "cellassign_celltype_annotation"
                          )

(p1|p2)/(p3|p4) /(p5|p6)

@@

#### Annotation versus `Seurat` clusters

Here we check wether one (or more) `Seurat` cluster do correspond to one label. 
This would allow us to be more confident in the annotation of immune or endothelial cells for example.

@@

#### Comparison of the labels from the two fetal references

The use of the right reference is crucial for label transfer and annotation. 
It is recommended that the cell types in the reference is representative to the cell types to be annotated in the query.

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)]. 
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

In the previous step of the analysis, we selected two fetal references containing kidney samples and transfer the labels.

Here we aim to understand the overlap and differences between the two references.

##### Human fetal kidney atlas Stewart et al.

Stewart et al. created a [human fetal kidney atlas](https://www.kidneycellatlas.org/) [[3](https://www.science.org/doi/10.1126/science.aat5031)].

This reference contains only fetal kidney cells and has been precisely annotated by kidney experts.

##### Azimuth Human fetal atlas Cao et al.

`Azimuth` also provides a human fetal atlas as a reference [[4](https://www.science.org/doi/10.1126/science.aba7721)].

The data can be found on [Zenodo](https://zenodo.org/records/4738021#.YJIW4C2ZNQI).

The reference contain cells from 15 organs including kidney from fetal samples.

We were first nicely surprised that most of the cells in the Wilms tumor dataset mapped to "kidney" organ (see previous report `02a_fetal_all_reference_Cao_{sample_id}.html`).

##### first level annotation

```{r fig.width=20, fig.height=8, out.width='100%'}
p1 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.compartment",
                            first_group = "fetal_full_predicted.annotation.l1"

                            )
p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "fetal_full_predicted.annotation.l1"

                            )
p1|p2
```

##### second level annotation

```{r fig.width=20, fig.height=8, out.width='100%'}
p2 <- do_Table_Heatmap( srat,
                            last_group = "fetal_kidney_predicted.cell_type",
                            first_group = "fetal_full_predicted.annotation.l2"

                            )
p2
```
